---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
```  
Note: knitr and ggplot2 warn about being built using R-3.1.2.  
I started using R-3.1.2 with R-studio and got errors so reverted to 3.1.1.

## Loading and preprocessing the data
  
Get and unpack the .csv file and check size 
```{r inputdata, results='asis'}
zipfile <- "./activity.zip"
csvfile <- unzip(zipfile, list=FALSE)
dataframe <- read.csv(csvfile, header=TRUE) 
d <- dim(dataframe) 
```  
The data has  `r d[1]`  observations and `r d[2]` variables

## What is mean total number of steps taken per day?


Method: Sum the steps for each day separately and then take the mean of these sums and the median of the sums

```{r Q1mean, results='asis'}
daysums <- aggregate(dataframe$steps,  by = list(dataframe$date),  sum)
names(daysums) <- c("date", "steps")
daymean1 <- mean(daysums$steps, na.rm=TRUE)
daymedian1 <- median(daysums$steps, na.rm=TRUE)
paste("The day step counts have mean = ", round(daymean1, 2),
        " and median = ", round(daymedian1, 2))
```  
  
Histogram of the frequencies of daily steps,
ie how many times each step count occured
```{r Q1plot, fig.height=3, fig.width=5}
hist(daysums$steps, 20, main="Frequency of the steps per day", 
     xlab="Number of steps in  day")
```

## What is the average daily activity pattern?

Method: Group the observations by each 5-minute interval and find the mean of each group
```{r Q2mean, results='asis'}
intmeans <- aggregate(dataframe$steps,  by = list(dataframe$interval),
                      FUN=function(x) mean(x, na.rm=TRUE))
names(intmeans) <- c("interval", "steps")
```
  
Find the 5-minute interval containing the maximum mean
```{r Q2max, results='asis'}
rowindex <- which(intmeans$steps == max(intmeans$steps))
maxmeans <- intmeans[rowindex,]
```    
Row `r rowindex` represents 5-minute interval  `r maxmeans$interval` which contains the maximum mean = `r round(maxmeans$steps,2)`  
`  
Time Series plot of the means of 5-minute interval steps on average across all the days,

```{r Q2plot, fig.height=3, fig.width=5}
plot(intmeans$steps, type="l",  
     main="Average steps per interval across all days",
     xlab=paste(nrow(intmeans)," 5-minute intervals per day"),
     ylab="mean steps/interval")
```


## Imputing missing values

There are `r nrow(dataframe[is.na(dataframe$steps),])` occurrences containing steps = NA  
  
Method: Replace the NA steps by the means from the corresponding intervals
```{r meansForNA, results='asis'}
# get means for intervals, (including NAs)
im <- aggregate(dataframe$steps,  by = list(dataframe$interval), 
                FUN=function(x) mean(x, na.rm=TRUE))
names(im) <- c("interval", "steps")

# Copy the dataframe and find the rows with NA
df <- dataframe
change <- which(is.na(df$steps))       # gets indices of NA rows

# for each NA row find the corresponding steps mean for its interval
#    # brute force method!
for (ic in change){                    # intervals with NA         
  # get the matching interval and store the mean    
  findim <- which(df$interval[ic] == im$interval)
  df$steps[ic] <- im$steps[findim]
}
```  
There are now `r nrow(df[is.na(df$steps),])` NA rows.  

Get the mean and median of the total steps for each day
```{r Q3mean, results='asis'}
daysums <- aggregate(df$steps,  by = list(df$date),  sum)
daymean2 <- mean(daysums$x, na.rm=TRUE)
daymedian2 <- median(daysums$x, na.rm=TRUE)
paste("The total steps for the days have mean  = ", 
    round(daymean2, 2), "  and median  = ", round(daymedian2, 2))
```  

The increase in the mean and median, due to filling in NA, is `r daymean2 - daymean1` and `r round(daymedian2 - daymedian1,2)` respectively.  

Histogram of the frequencies of total steps for each day  
ie how many times each step count occured
```{r Q3plot, fig.height=3, fig.width=5}
hist(daysums$x, 20, main="Occurrences of the steps taken per day", 
     sub="(each NA replaced by its interval mean)",
     xlab="Number of steps in  day")
```  


## Are there differences in activity patterns between weekdays and weekends?
Add a column named "day" identifying weekdays/weekends
```{r Weekends, results='asis'} 
df$day <- factor( ifelse( weekdays(as.POSIXlt(df$date)) %in% as.list(c("Saturday","Sunday")), "weekend", "weekday"))

# group steps by interval and day and get the group means
wdays <- aggregate(df$steps,  by = list(df$day, df$interval),  mean) 
names(wdays) <- c("day", "interval", "steps")
```  
Compare the weekday to weekend step activity for each interval  
```{r plot2compare, fig.height=4, fig.width=5  }
library(ggplot2)
ggplot(wdays, aes(interval, steps)) +
  geom_line() +
  ggtitle("Comparison of weekday to weekend activity")+
  ylab("number of steps")+
  xlab("5-minute intervals")+
  facet_wrap(~day, nrow=2)
```  

## End of Reproducible Results Assignment 1  


